Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-28 SIR_manuscript.tex
addressing the comments

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ in the following form.
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where the term $\frac{\beta}{\gamma}$ is the classical basic reproduction number for a SIR model without testing and isolation \citep{keeling2011modeling}, and the positive quantity $\Delta$ is the effectiveness of control due to testing and isolation processes defined as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C\,N_0}\big(C_1\,S\_u^*+(C_2(1-\theta\_w)+C\,\theta\_w)\,S\_n^*\big),
\end{equation}
\bmb{super-fussy, but: I don't think we should introduce explicit thinspaces everywhere; why not let
  TeX do the typesetting itself? Do we really prefer more compact equations, and if so is there a package
  that will just change the appropriate settings globally?}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} =& (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2\,(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2\,p_I\,\hat F_I\,\theta\_c.
\end{align}
\bmb{I don't like the alignment of the equations: too little space after =. I realize that's incredibly picky.}
Further details of derivation of $\Rnum$ are provided in appendix, Sec.\,\ref{app:R0}. 
The explicit formula of $\Rnum$ enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (Sec.\,\ref{app:rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Rnum$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Rnum$ and $\omega$. \bmb{The previous version of this implied that the restriction was $\rho \ll N_0$; that seems very odd since $\rho$ has units of time$^{-1}$. I think $\rho \ll 1$ is the correct statement (i.e. $\rho$ is already a \emph{per capita} measure), but please check. If I am wrong and $\rho$ is expressed in absolute units of tests/day, then the range quoted in the caption for \fref{pan} is weird \ldots}

The analytical calculation of the next-generation matrix and simplification of the $\Rnum$ expression was carried out in Maple \citep{maple14}. We used R \citep{r} for numerical solutions, in particular for plotting the contours of $\Delta$ (Eq.~\eqref{NOTSURE}) \bmb{not sure what the right ref. is. This was previously ``eq:tel4'', which doesn't exist. Should it have been ``eq:del4''?} over a range of parameters.
We computed the values and contours of Delta at both low (\fref{pan}) and high (\fref{pan2}) testing intensities, and for both random testing ($w_S=w_I=w_R=1$) and targeted testing ($w_S=0.3$; $w_I=w_R=1$).

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is $\approx 1.017$, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing. \bmb{how did we choose this particular value?}
  \bmb{JD wonders whether we really need the ``critical'' $\Rnum=1$ contour or not; any contour will do to illustrate a threshold level of testing intensity, and the fact that $\Rnum=1$ falls in the middle of our range is a bit arbitrary (i.e. the fact that testing can control the epidemic depends on the baseline $\Rnum$ value we have chosen). (I know it was my idea in the first place, but maybe we could dispense with it?)}

>>>>

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ in the following form.
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where the term $\frac{\beta}{\gamma}$ is the classical basic reproduction number for a SIR model without testing and isolation \citep{keeling2011modeling}, and the positive quantity $\Delta$ is the effectiveness of control due to testing and isolation processes defined as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
\bmb{super-fussy, but: I don't think we should introduce explicit thinspaces everywhere; why not let
  TeX do the typesetting itself? Do we really prefer more compact equations, and if so is there a package
  that will just change the appropriate settings globally?}\ali{are we talking about \,? not sure about the package!}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
Further details of derivation of $\Rnum$ are provided in appendix, Sec.\,\ref{app:R0}. 
The explicit formula of $\Rnum$ enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (Sec.\,\ref{app:rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Rnum$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Rnum$ and $\omega$.

The analytical calculation of the next-generation matrix and simplification of the $\Rnum$ expression was carried out in Maple \citep{maple14}. We used R \citep{r} for numerical solutions, in particular for plotting the contours of $\Delta$ (Eq.~\eqref{eq:del4a}) over a range of parameters.
We computed the values and contours of Delta at both low (\fref{pan}) and high (\fref{pan2}) testing intensities, and for both random testing ($w_S=w_I=w_R=1$) and targeted testing ($w_S=0.3$; $w_I=w_R=1$).

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is $\approx 1.017$, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing. \bmb{how did we choose this particular value?}\ali{I was thinking of ``critical'' $\Rnum=1$ and wanted to have the baseline in the plot.}
  \bmb{JD wonders whether we really need the ``critical'' $\Rnum=1$ contour or not; any contour will do to illustrate a threshold level of testing intensity, and the fact that $\Rnum=1$ falls in the middle of our range is a bit arbitrary (i.e. the fact that testing can control the epidemic depends on the baseline $\Rnum$ value we have chosen). (I know it was my idea in the first place, but maybe we could dispense with it?)}\ali{agree with the point, the plots will be modified.}
